---
title: "Clustering and risk scenarios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clustering and risk scenarios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "figures/3-",
  out.width = "100%",
  echo     = TRUE,
  message  = FALSE,
  warning  = FALSE
)

set.seed(42)

# Set BCUBED_DOCS_BUILD=true in CI / docs conversion to skip heavy chunks if needed.
BCUBED_DOCS_BUILD = identical(Sys.getenv("BCUBED_DOCS_BUILD"), "true")
```


```{r read_files, include=FALSE}
# At the start of the next vignette, explicitly load files from previous vignette:
inputs = readRDS("_data/inputs_vignettes.rds")

fit = inputs$fit
site_df = inputs$site_df
fitness_df = inputs$fitness_df
df_inv = inputs$df_inv
traits_inv = inputs$traits_inv
```

---

## Install and load `invasimapr` and other core packages

Install from GitHub (recommended for development) or CRAN (if available) as before. 

```{r install-invasimapr, include=FALSE}
# install.packages("remotes")
# remotes::install_github("b-cubed-eu/invasimapr")

# If invasimapr is on CRAN:
# install.packages("invasimapr")

library(invasimapr)
library(ggplot2)
library(dplyr)   # <-- needed for across(), left_join()

deps = c(
  # Core workflow
  "dplyr","tidyr","purrr","tibble","stringr",
  "ggplot2",
  # Modelling / stats
  "glmmTMB","performance","Matrix","glmnet","MASS","vegan","cluster","matrixStats",
  # Spatial
  "sf",
  # Viz helpers used in later steps
  "viridis","pheatmap","factoextra"
)

is_installed = vapply(deps, requireNamespace, logical(1), quietly = TRUE)
missing = deps[!is_installed]

cat("### Dependency check\n\n")
if (length(missing) == 0) {
  cat("- All recommended packages are available.\n")
} else {
  cat("- Missing packages (install before running the full workflow):\n")
  cat(paste0("  - ", missing, collapse = "\n"), "\n")
}
```

---

## Clustering and risk scenarios

After computing invasion fitness $\lambda_{is}$ (Sections 6-7), we can mine the $\text{invader} \times \text{site}$ surface for **recurring profiles**. Clustering reveals groups of invaders with similar spatial success and groups of sites with similar susceptibility, enabling practical **risk categories** and **scenario planning**.

:hourglass_flowing_sand: **Step-by-step** this section:

1.  We reshape the long fitness table into a matrix $\Lambda$ with rows = invaders and columns = sites;
2.  scale and cluster rows and columns (Ward's method) to expose structure;
3.  assign **discrete cluster labels** to invaders and sites for downstream mapping and summaries; and
4.  visualise the joint structure with a clustered heatmap.

---

### Hierarchical clustering of invaders and sites

We treat each invader as a vector of fitness values across sites and each site as a vector across invaders. Hierarchical clustering groups **invaders** into "broad-spectrum" vs. "specialists" and **sites** into "open" vs. "resistant" community types. We scale rows/columns to neutralise magnitude, compute Euclidean distances, and apply Ward's linkage. The number of clusters $k$ can be chosen by silhouette/gap heuristics or fixed for interpretability.

:hourglass_flowing_sand: **Step-by-step** this section: 1. Build the $\lambda_{is}$ matrix &#x2794; 2. Clean &#x2794; 3. Scale rows/cols &#x2794; 4. Compute Euclidean distances &#x2794; 5. Cluster with Ward’s linkage &#x2794; 6. Choose ($k$) (silhouette when available; else a small default) &#x2794; 7. Attach cluster labels back to data.

---

#### Setup helper functions

```{r cluster-setup}
# Optional for k selection via silhouette:
has_fviz = requireNamespace("factoextra", quietly = TRUE)

# Helpers --------------------------------------------------------------
safe_scale = function(M) {
  if (is.null(dim(M))) return(matrix(M, nrow = 1L))
  sds = apply(M, 2, sd, na.rm = TRUE)
  z = scale(M, center = TRUE, scale = sds)
  z[is.na(z)] = 0  # handles sd=0 columns
  z
}

safe_kmax = function(X, kmax = 10L) {
  n = nrow(X); if (is.null(n) || n < 3L) return(NA_integer_)
  min(kmax, n - 1L)  # k in [2, n-1]
}

choose_k_silhouette = function(X, kmax = 10L, nstart = 25L) {
  kmax_safe = safe_kmax(X, kmax)
  if (is.na(kmax_safe) || !has_fviz) return(NA_integer_)
  gg = factoextra::fviz_nbclust(
    as.data.frame(X), FUNcluster = kmeans, method = "silhouette",
    k.max = kmax_safe, nstart = nstart
  )
  d = gg$data
  if (!is.null(d) && all(c("clusters","y") %in% names(d))) d$clusters[which.max(d$y)] else NA_integer_
}
```

---

#### Assemble the invader × site fitness matrix ($\lambda$)

We pivot long results (one row per site×invader) to a wide matrix with **rows = invaders**, **cols = sites**. If multiple draws exist, we take the mean per cell. Then drop rows/columns that are entirely `NA` (or all missing after earlier filters). If too few remain, we bail out gracefully.

```{r cluster-mat}
lambda_mat = fit$summary$establish_long |>
  dplyr::select(invader, site, lambda) |>
  tidyr::pivot_wider(
    id_cols = invader,
    names_from = site,
    values_from = lambda,
    values_fill = 0,
    values_fn = list(lambda = ~ mean(.x, na.rm = TRUE))
  ) |>
  tibble::column_to_rownames("invader") |>
  as.matrix()

dim(lambda_mat)     # rows=invaders, cols=sites

lambda_mat_noNA = lambda_mat[
  rowSums(is.na(lambda_mat)) < ncol(lambda_mat),
  colSums(is.na(lambda_mat)) < nrow(lambda_mat),
  drop = FALSE
]

if (nrow(lambda_mat_noNA) < 2L || ncol(lambda_mat_noNA) < 2L) {
  warning("Too few invaders or sites for clustering; assigning NA clusters.")
}
dim(lambda_mat_noNA)     # rows=invaders, cols=sites

```

---

#### Cluster **sites** (columns as observations)

We scale invader dimensions (columns) to neutralize magnitude and cluster sites via `Ward.D2`.

```{r cluster-sites}
if (nrow(lambda_mat_noNA) >= 2L && ncol(lambda_mat_noNA) >= 2L) {

  X_sites = t(safe_scale(lambda_mat_noNA))  # rows = sites, cols = invaders
  keep_sites = apply(X_sites, 1, sd) > 0
  if (any(!keep_sites)) X_sites = X_sites[keep_sites, , drop = FALSE]

  site_dist  = dist(X_sites)
  site_clust = hclust(site_dist, method = "ward.D2")

  k_sites = choose_k_silhouette(X_sites, kmax = 10L)
  if (is.na(k_sites)) k_sites = max(2L, min(5L, nrow(X_sites)))

  site_groups_ok = cutree(site_clust, k = k_sites)  # named vector (kept sites)

  # Reinsert dropped sites (if any) as NA; preserve original site order
  site_groups = setNames(rep(NA_integer_, ncol(lambda_mat_noNA)),
                          colnames(lambda_mat_noNA))
  site_groups[names(site_groups_ok)] = site_groups_ok
}
```

---

#### Cluster **invaders** (rows as observations)

Same idea, now with rows (invaders) as observations.

```{r cluster-invaders}
if (nrow(lambda_mat_noNA) >= 2L && ncol(lambda_mat_noNA) >= 2L) {

  X_inv = safe_scale(lambda_mat_noNA)       # rows = invaders, cols = sites
  keep_inv = apply(X_inv, 1, sd) > 0
  if (any(!keep_inv)) X_inv = X_inv[keep_inv, , drop = FALSE]

  inv_dist  = dist(X_inv)
  inv_clust = hclust(inv_dist, method = "ward.D2")

  k_inv = choose_k_silhouette(X_inv, kmax = 10L)
  if (is.na(k_inv)) k_inv = max(2L, min(5L, nrow(X_inv)))

  inv_groups_ok = cutree(inv_clust, k = k_inv)  # named vector (kept invaders)

  inv_groups = setNames(rep(NA_integer_, nrow(lambda_mat_noNA)),
                         rownames(lambda_mat_noNA))
  inv_groups[names(inv_groups_ok)] = inv_groups_ok
}
```

---

#### Attach cluster labels for downstream use

We create (or update) a `df_inv` table with one row per `(site, invader)` and two factor columns: `site_cluster`, `invader_cluster`. Any dropped entities become `NA`.

```{r cluster-labels}
# Ensure df_inv exists
if (!exists("df_inv")) {
  df_inv = fit$summary$establish_long |>
    dplyr::select(site, invader) |>
    dplyr::distinct() |>
    dplyr::mutate(across(c(site, invader), as.character))
}

# Sanity checks: names should align with the clustered matrix
if (!all(df_inv$site %in% names(site_groups))) {
  warning("Some sites in df_inv not present in clustered matrix; setting NA.")
}
if (!all(df_inv$invader %in% names(inv_groups))) {
  warning("Some invaders in df_inv not present in clustered matrix; setting NA.")
}

df_inv = df_inv |>
  dplyr::mutate(
    site_cluster    = factor(site_groups[site],
                             levels = sort(unique(site_groups))),
    invader_cluster = factor(inv_groups[invader],
                             levels = sort(unique(inv_groups)))
  )

head(df_inv)
```

---

:information_source: **Why this matters**: Cluster labels compress thousands of $\lambda_{is}$ cells into a handful of **profiles** that managers and analysts can reason about.

:warning: **Checks/tips**: 
* Remove rows/cols with only `NA` before clustering; clustering needs variation.
* Scaling neutralizes magnitude so distance reflects **shape** of response.
* If silhouette selection isn’t available, we default to a small ($k$) (2–5) for interpretability.
* Store `site_cluster` and `invader_cluster` on your canonical tables for downstream stratified summaries, e.g. site-type vs invader-type risk maps.

---

#### Quick diagnostic plots

:bar_chart: **Clustered** $\lambda$ heatmap summarizes the joint structure of invasion fitness ($\lambda$) across invaders (rows) and sites (columns).

```{r cluster-sites_dendrogram, fig.alt="Site dendrogram"}
# Sites dendrogram
if (exists("site_clust")) plot(site_clust, main = "Sites (Ward.D2)", xlab = "", sub = "")
```

```{r cluster-invaders_dendrogram, fig.alt="Invader dendrogram"}
# Invaders dendrogram
if (exists("inv_clust")) plot(inv_clust, main = "Invaders (Ward.D2)", xlab = "", sub = "")
```

> :chart_with_upwards_trend: **Figure 24**: Dendrograms, useful to eyeball structure and plausibility (k).


```{r fit-pheatmap, fig.alt="Clustered heatmap of invasion fitness"}
# Clustered heatmap: joint structure in $\lambda$ (invader × site)
pheatmap::pheatmap(
  lambda_mat_noNA,
  kmeans_k = 5,
  color = rev(viridis::viridis(50, option = "magma", direction = 1)),
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method        = "ward.D",
  fontsize_row = 8, fontsize_col = 8,
  main = "Clustered Invasion Fitness Matrix (Invader × Site)",
  angle_col = 45
)
```

> :chart_with_upwards_trend: **Figure 25**: The clustered heatmap summarizes the joint structure of invasion fitness ($\lambda$) across invaders (rows) and sites (columns): Colors encode $\lambda$ (yellow/white = higher fitness; purple/black = lower), and the dendrograms group invaders and sites by similarity in their $\lambda$ profiles (Ward clustering). Clear vertical bands of warmer colors indicate that site identity explains much of the variation, certain site clusters consistently offer higher fitness across many invaders, whereas horizontal blocks are weaker, implying more modest between-invader differences. The k-means split (k = 5) on the columns highlights a few distinct site regimes that alternate between high and low $\lambda$, producing contiguous stripes rather than scattered pixels. Some invader clusters show uniformly poor performance (cool rows) with only small pockets of opportunity, while others have broader tolerance with multiple site clusters showing elevated $\lambda$. **Overall, the figure reveals strong spatial structure in establishment potential, with invader effects present but secondary to site-level context.**

:sparkles: **Overall importance**: Clustering turns a dense prediction surface into a concise **vocabulary of profiles** for communication and planning.

:bulb: **Summary.** You now have **invader** and **site** cluster labels that can be mapped, ranked, and cross-tabulated with traits and environments.

------------------------------------------------------------------------

### Mapping site-level risk categories

To support spatial decisions, we translate site clusters into **ordered risk categories** (*very-high* → *very-low*) using mean $\lambda_{is}$ within each cluster, optionally constraining clusters to be geographically cohesive. In this section:

1. We use **ClustGeo** to blend similarity in $\Lambda$ profiles with geographic distance, 
2. **choose $k$ clusters**, and then 
3. **relabel** clusters by their mean site fitness into intuitive **risk categories**.
4. Finally, we **map categories** over the study region.

:information_source: **Why this matters**: The map highlights **where** openness to invasion concentrates and provides a stable spatial unit for monitoring and intervention.

:warning: **Checks/tips**: Ensure coordinate joins are correct; test a few $\alpha$ weights (0 = spatial only, 1 = fitness only); order categories by mean fitness for consistent legends.

:hourglass_flowing_sand: **Step-by-step**: 1. Assemble $\lambda$ &#x2794; 2. Prepare site coordinates &#x2794; 3. Build profile and spatial distances &#x2794; 4. Cluster with `ClustGeo` &#x2794; 5. Choose $k$ &#x2794; 6. Relabel clusters by mean $\lambda$ &#x2794; 7. Map

---

#### Setup (packages, inputs)

```{r risk-setup}
# Optional: spatial / clustering helpers
has_sf      = requireNamespace("sf", quietly = TRUE)
has_clgeo   = requireNamespace("ClustGeo", quietly = TRUE)

# Expected objects:
# - fit$summary$establish_long or fit$fitness$lambda_long with (site, invader, lambda)
# - fitness_df with site-level summaries incl. site coords (x,y) or an sf geometry
# - rsa (optional sf boundary for context)
```

---

#### Assemble a long table of ($\lambda$) (one row per site × invader)

Guard against duplicates by averaging within (site, invader) and drop fully-NA rows/columns for stability.

```{r risk-table}
lambda_long = if (!is.null(fit$fitness$lambda_long)) {
  fit$fitness$lambda_long
} else {
  fit$summary$establish_long
}

lambda_mean = lambda_long |>
  dplyr::select(site, invader, lambda) |>
  dplyr::mutate(across(c(site, invader), as.character)) |>
  dplyr::group_by(site, invader) |>
  dplyr::summarise(mean_lambda = mean(lambda, na.rm = TRUE), .groups = "drop")

# Drop fully-NA rows/columns for stability
lambda_mat = lambda_mean |>
  tidyr::pivot_wider(
    id_cols = invader, names_from = site, values_from = mean_lambda,
    values_fill = 0, values_fn = list(mean_lambda = ~ mean(.x, na.rm = TRUE))
  ) |>
  tibble::column_to_rownames("invader") |>
  as.matrix()
dim(lambda_mat)

lambda_mat_noNA = lambda_mat[
  rowSums(is.na(lambda_mat)) < ncol(lambda_mat),
  colSums(is.na(lambda_mat)) < nrow(lambda_mat),
  drop = FALSE
]

stopifnot(ncol(lambda_mat_noNA) >= 2L, nrow(lambda_mat_noNA) >= 2L)
dim(lambda_mat_noNA)
```

---

#### Prepare site coordinates (x,y)

Accepts `fitness_df` either with numeric `x,y` columns *or* as `sf` with point geometry.

```{r risk-sites}
# Extract site ids to be clustered (columns of lambda_mat_noNA)
site_ids = colnames(lambda_mat_noNA)

# Get coordinates for these sites
if (has_sf && inherits(fitness_df, "sf")) {
  # If sf, derive lon/lat (or projected) from geometry
  xy = sf::st_coordinates(sf::st_centroid(fitness_df))
  coords_df = cbind(st_drop_geometry(fitness_df), x = xy[,1], y = xy[,2])
} else {
  coords_df = fitness_df
}

stopifnot(all(c("site","x","y") %in% names(coords_df)))
site_coords = coords_df |>
  dplyr::mutate(site = as.character(site)) |>
  dplyr::filter(site %in% site_ids) |>
  dplyr::select(site, x, y)

# Reorder to match site_ids; keep only complete cases
site_coords = site_coords[match(site_ids, site_coords$site), ]
ok = stats::complete.cases(site_coords$x, site_coords$y)
stopifnot(any(ok))  # need at least some sites with coords
```

---

#### Build distances: profile (D1) + geographic (D0)

Scale site profiles (rows) to neutralize magnitude; then compute Euclidean distances.

```{r risk-distance}
# X = sites × invaders (scaled across invaders)
scale_cols = function(M) {
  sds = apply(M, 2, sd, na.rm = TRUE)
  Z   = scale(M, center = TRUE, scale = sds); Z[is.na(Z)] = 0; Z
}

X  = scale_cols(t(lambda_mat_noNA[, ok, drop = FALSE]))  # rows = sites(ok)
D1 = dist(X)                                            # profile distance

D0 = dist(as.matrix(site_coords$x[ok]) |> cbind(site_coords$y[ok]))
# Equivalent: dist(cbind(site_coords$x[ok], site_coords$y[ok]))
```

---

#### Run ClustGeo (blend profile + spatial cohesion)

Tune ($\alpha \in$ [0,1]): 0 = spatial only; 1 = profile only. *(Optional)* try a few ($\alpha$) values and inspect `inertia` to pick a trade-off.

```{r risk-ClustGeo}
if (!has_clgeo) stop("ClustGeo not installed. install.packages('ClustGeo')")

alpha = 0.3   # adjust: higher = prioritize lambda profiles; lower = more spatial cohesion
tree  = ClustGeo::hclustgeo(D0, D1, alpha = alpha)

# Choose k (fixed or heuristic). For spatial categories, a small k is typical.
k = 5
site_groups_ok = stats::cutree(tree, k = k)  # names are sites with ok coords

# Reinsert NAs for sites lacking coords; order = site_ids
site_groups = setNames(rep(NA_integer_, length(site_ids)), site_ids)
site_groups[names(site_groups_ok)] = site_groups_ok
```

*Optional: alpha tuning curve*

```{r risk-alpha, fig.alt="Effect of `ClustGeo` mixing parameter ($\alpha$) on cluster compactness"}
# Within-cluster sum of squares from a distance matrix
# Works for Euclidean distances (e.g., dist() on scaled data or coordinates).
# W(C) = 0.5 / n_C * sum_{i,j in C} d(i,j)^2  ;  W = sum_C W(C)

withinss_dist = function(D, groups) {
  if (inherits(D, "dist")) D = as.matrix(D)
  g = as.factor(groups)
  idx = split(seq_len(nrow(D)), g)
  w = vapply(idx, function(I) {
    n = length(I)
    if (n <= 1L) return(0)
    Dij2 = D[I, I, drop = FALSE]^2
    sum(Dij2) / (2 * n)
  }, numeric(1))
  sum(w)
}

# Total sum of squares from a distance matrix: T = 0.5/n * sum_{i,j} d(i,j)^2
totss_dist = function(D) {
  if (inherits(D, "dist")) D = as.matrix(D)
  n = nrow(D)
  if (n <= 1L) return(0)
  sum(D^2) / (2 * n)
}

# Evaluate a grid of alpha to visualize within-cluster inertia trade-off
alphas = seq(0, 1, by = 0.1)

# Precompute totals for normalization (optional)
T1 = totss_dist(D1)   # profile total inertia
T0 = totss_dist(D0)   # spatial total inertia

alpha_curve = lapply(alphas, function(a) {
  tr = ClustGeo::hclustgeo(D0, D1, alpha = a)
  gr = cutree(tr, k = k)

  W1 = withinss_dist(D1, gr)         # within-cluster profile inertia
  W0 = withinss_dist(D0, gr)         # within-cluster spatial inertia
  data.frame(alpha = a,
             W1 = W1,  W1_rel = if (T1 > 0) W1/T1 else NA_real_,
             W0 = W0,  W0_rel = if (T0 > 0) W0/T0 else NA_real_)
})
alpha_curve = do.call(rbind, alpha_curve)

# Plot (relative values 0–1 are easiest to compare)
ggplot2::ggplot(alpha_curve, ggplot2::aes(alpha)) +
  ggplot2::geom_line(ggplot2::aes(y = W1_rel), linewidth = 0.8) +
  ggplot2::geom_point(ggplot2::aes(y = W1_rel), size = 1.2) +
  ggplot2::geom_line(ggplot2::aes(y = W0_rel), linetype = 2) +
  ggplot2::geom_point(ggplot2::aes(y = W0_rel), size = 1.2, shape = 1) +
  ggplot2::labs(y = "Relative within-cluster inertia",
       title = "Alpha trade-off: profile (solid) vs spatial (dashed)",
       subtitle = paste("k =", k)) +
  ggplot2::theme_minimal(11)

```

> :chart_with_upwards_trend: **Figure 26**: Trade-off curve showing how the choice of the `ClustGeo` mixing parameter ($\alpha$) affects clustering compactness in trait-profile space (solid line) versus geographic space (dashed line). At ($\alpha=0$), clusters are maximally cohesive geographically but poorly capture profile similarity; at ($\alpha=1$), the reverse holds. The curve suggests that intermediate values ($(\alpha \approx 0.3!-!0.6)$) provide a reasonable balance, with modest loss of profile compactness but large gains in spatial cohesion.

---

#### Order clusters by mean site fitness → ordered **risk categories**

Compute a site-level mean ($\lambda$); order clusters by descending mean; map to labels.

```{r risk-labels}
# Site-level mean lambda
site_lambda_mean = lambda_mean |>
  dplyr::group_by(site) |>
  dplyr::summarise(lambda_mean = mean(mean_lambda, na.rm = TRUE), .groups = "drop")

# Rank clusters by mean site lambda
tmp = data.frame(site = names(site_groups), cluster = site_groups) |>
  dplyr::left_join(site_lambda_mean, by = "site") |>
  dplyr::group_by(cluster) |>
  dplyr::summarise(mu = mean(lambda_mean, na.rm = TRUE), .groups = "drop") |>
  dplyr::arrange(desc(mu))

risk_labels = c("very-high","high","medium","low","very-low")[seq_len(nrow(tmp))]
remap = setNames(risk_labels, tmp$cluster)

# Final site table with ordered categories
site_sum = site_coords |>
  dplyr::transmute(site, x, y,
            site_cluster = remap[as.character(site_groups[site])],
            site_category = factor(site_cluster, levels = risk_labels))
```

---

#### Map the categories

Use either a gridded backdrop or your `sf` boundary. If `fitness_df` is gridded, tiles work; if polygons, join on `site` and plot geometry.

```{r risk-map, fig.alt="Spatial categories of invasion risk"}
p = ggplot2::ggplot(site_sum, ggplot2::aes(x = x, y = y, fill = site_category)) +
  ggplot2::geom_tile(color = NA) +
  ggplot2::labs(title = "Spatial invasion risk (ClustGeo-constrained categories)",
       x = "Longitude", y = "Latitude", fill = "Site invasibility") +
  ggplot2::scale_fill_brewer(palette = "RdYlBu", direction = 1, drop = FALSE) +
  ggplot2::theme_minimal(12) +
  ggplot2::theme(panel.grid = ggplot2::element_blank(),
        legend.position = "right")

# Add boundary if available
if (exists("rsa") && has_sf && inherits(rsa, "sf")) {
  p = p + ggplot2::geom_sf(data = rsa, inherit.aes = FALSE, fill = NA, color = "black", size = 0.5)
}
p
```

> :chart_with_upwards_trend: **Figure 27**: The map depicts **spatial categories of invasion risk** derived from a ClustGeo clustering that jointly considers site-level invasion fitness ($\lambda$ profiles across invaders) and geographic proximity. Each grid cell is assigned to one of five ordered categories, very high, high, medium, low, and very low invasibility, based on the mean establishment potential of invaders at that location.

:bar_chart: Two patterns emerge: First, invasion risk is **spatially structured**, with contiguous clusters of high-risk sites (red) concentrated in the southern and central regions, and very-high categories also evident in the northeast. These regions combine favorable abiotic conditions with invader-compatible resident communities. Second, low and very-low risk sites (blue shades) form coherent zones, often in the western and coastal areas, suggesting environmental filtering or strong resident resistance. Because the categories are **constrained by geography**, the map highlights regional-scale patterns rather than isolated hotspots. This makes it useful for decision-making, since it aligns biological risk signals with practical spatial units for monitoring and intervention. **In short, the figure reveals where invasion pressure is expected to be greatest, where resistance is likely strongest, and how these patterns are distributed across the South African landscape.**

---

### Cluster-wise summaries for reporting

Summarize mean/SD/CV of ($\lambda$) **by risk category**, and across **(site × invader)** cells.

```{r risk-sumaries}
lambda_with_cat = lambda_mean |>
  dplyr::left_join(dplyr::select(site_sum, site, site_category), by = "site")

summ_sitecat = lambda_with_cat |>
  dplyr::group_by(site_category) |>
  dplyr::summarise(
    n_cells = dplyr::n(),
    mean_lambda = mean(mean_lambda, na.rm = TRUE),
    sd_lambda   = sd(mean_lambda, na.rm = TRUE),
    cv_lambda   = sd_lambda / mean_lambda,
    q10 = quantile(mean_lambda, 0.10, na.rm = TRUE),
    q50 = quantile(mean_lambda, 0.50, na.rm = TRUE),
    q90 = quantile(mean_lambda, 0.90, na.rm = TRUE),
    .groups = "drop"
  )

summ_sitecat
```

*Quick barplot of category means:*

```{r risk-barplot, fig.alt="Mean invasion fitness"}
ggplot2::ggplot(summ_sitecat,
       ggplot2::aes(x = site_category, y = mean_lambda, fill = site_category)) +
  ggplot2::geom_col(show.legend = FALSE) +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = mean_lambda - 1.96*sd_lambda/sqrt(n_cells),
                    ymax = mean_lambda + 1.96*sd_lambda/sqrt(n_cells)),
                width = 0.15) +
  ggplot2::labs(x = "Risk category", y = expression("Mean " * lambda),
       title = expression("Category-wise summaries of " * lambda)) +
  ggplot2::theme_minimal(11)
```

> :chart_with_upwards_trend: **Figure 28**: Category-wise mean invasion fitness ($\lambda$) across ordered site risk categories (*very-high* to *very-low*). Sites assigned to the *very-high* category exhibit near-zero or slightly positive mean fitness values, indicating conditions most favorable to invader establishment. In contrast, *low* and *very-low* categories show strongly negative mean fitness, reflecting resistant community conditions. The monotonic decline confirms that the clustering and relabeling successfully captured a gradient of invasion risk.

---

:sparkles: **Overall importance**: Clustering + mapping yields actionable **spatial units** and **invader types** for prioritising surveillance and control.

:bulb: **Summary.** You now have spatially coherent **site categories** and interpretable **invader categories** that align with the full fitness surface.

:warning: **Checks/tips**: 
* **Coordinates:** confirm site ↔ coord join; if using `sf`, verify CRS.
* **Alpha:** test a few ($\alpha$) (e.g., 0, 0.3, 0.6, 1) and inspect inertia to trade off cohesion vs profile similarity.
* **k:** small ($k$) (3-6) aids communication; fix ($k$) for comparability across runs.
* **Ordering:** always order labels by **descending** mean ($\lambda$) to keep legends consistent.
* **Missing coords:** sites without coordinates get `NA` clusters; handle explicitly if mapping is required.


```{r save-state, include=FALSE}
# # Save state for downstream vignettes
# # Each vignette runs in a clean R session. We therefore **persist state explicitly**.
# dir.create("_data", showWarnings = FALSE, recursive = TRUE)
# 
# saveRDS(
#   list(
#     fit = fit,
#     site_df = site_df,
#     fitness_df = fitness_df,
#     lambda_long = lambda_long,
#     df_inv = df_inv,
#     traits_inv = traits_inv
#   ),
#   # file = "_data/fit_cluster_inputs.rds",
#   file = "_data/inputs_3.rds",
#   compress = "xz"
# )
```

---

## Session information

```{r session-info}
sessionInfo()
```
